Making ice sheet models \\ scale properly

### 2023-02-10 talk/slides.tex
sweep through and simplify/clarify/shorten

\newcommand{\bb}{\mathbf{b}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}

@@

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$  is defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice!
    \end{itemize}
\item<3> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which (solution) velocity $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

@@

\medskip
\item conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] \emph{in} the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and \emph{on} the ice surfaces:}
\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}

@@

\bigskip
\item<2> {\footnotesize for more on this perspective see Bueler (2021)}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \only<1>{\alert{nonlinear complementarity problem} (NCP)} \only<2>{nonlinear complementarity problem (NCP) coupled to Stokes}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t)$}}
\end{align*}

@@

\begin{itemize}
\item for this slide, forget complementarity and boundary conditions
\item result is SKE coupled to Stokes:
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

\item<2-> only the first of these 5 equations has a time derivative
    \begin{itemize}
    \item[$\circ$] because ice is slow and incompressible
    \end{itemize}
\item<3> the problem is a \alert{differential algebraic equation} (DAE), an extremely \alert{stiff}, system:
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item<3>[$\circ$] but in $\infty$ dimensions (PDAE?) and subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying the well-posedness or regularity of this basic ice sheet model
    \begin{itemize}
    \item[$\circ$] but most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model?
    \item[$\circ$] well-posedness of shallow approximations \emph{has} been considered
    \end{itemize}

\medskip
\item<2> numerical modelers tend to think of the Stokes problem separately from surface evolution
    \begin{itemize}
    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is taken for granted
    \end{itemize}
\item<2> ice sheet geometry evolution is addressed with minimal awareness of complementarity
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item thickness transport form helps expose evolution character
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item<2-> the \alert{mass continuity equation} follows from the SKE and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<3-> \emph{question:} is this really an advection equation?
\item<3->[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H,$$
and ice is slow (DAE) so there are no characteristics
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item the diffusion schema is only literal in the lubrication approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item but the fact that \emph{ice flows downhill} has \emph{time-stepping stability consequences}
    \begin{itemize}
    \item[$\circ$] regardless of your advective schema preference
    \end{itemize}
\item in any case, both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
\end{itemize}
\end{frame}

\begin{frame}{shallow ice sheet models}

\begin{itemize}
\item the basic ice sheet model (NCP coupled to Stokes) is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}

\smallskip
\item current-technology models use various shallow simplifications
    \begin{itemize}
    \item[$\circ$] justification: ``Stokes is too expensive at scale''
    \end{itemize}
\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)

\smallskip
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item in the shallow ice approximation (SIA) version of the NCP,
{\small
  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
}
the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$

@@

\medskip
\item<2> $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item<2>[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item<2>[$\circ$] \emph{local} in surface and bed topography
    \item<2>[$\circ$] existence is known for this NCP problem (Jouvet \& Bueler, 2012)
         \begin{itemize}
         \item[$\vartriangleright$] once written as a \alert{variational inequality} weak form
         \end{itemize}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item from now on, let's try to avoid shallowness approximations
\item then the basic ice sheet model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
{\small
\begin{equation*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
\end{equation*}
}
\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the SIA velocity responds only underneath the surface perturbation}
\end{itemize}

@@

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item \emph{caution}: for ice sheet models, $\bU(t,x,y)$ and $D(t,x,y)>0$ vary in space and time, from nonlinear coupling
\item nonetheless, let us recall some traditional numerical analysis \dots
}
\only<2>{
\item \alert{explicit} time stepping is common for \alert{advections}
\item for example, forward Euler using spacing $h$ and time step $\Delta t$:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{(\bU H)_{j+1/2}^\ell - (\bU H)_{j-1/2}^\ell}{h} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with CFL maximum time step
        $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
    \item[$\circ$] need flux approximations: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \end{itemize}
}
\only<3>{
\item \alert{explicit} time stepping for \alert{diffusions} is best avoided
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell} + s_{j}^{\ell}) - D_{j-\frac{1}{2}} (s_{j}^{\ell} + s_{j-1}^{\ell})}{h^2} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with maximum time step
        $$\Delta t \le \frac{h^2}{\max D} = O(h^2)$$
    \end{itemize}
}
\only<4>{
\item \alert{implicit} time stepping for \alert{diffusions} is recommended
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{h^2} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] unconditionally stable, but must solve equations at each step
    \item[$\circ$] alternative implicit schemes: Crank-Nicolson, BDF, \dots
    \end{itemize}
}
\end{itemize}
\end{minipage}
\end{frame}

\begin{frame}{time-stepping in current ice sheet models}

\begin{itemize}
\item most researchers believe the advection schema
\item current-technology models mostly use explicit time stepping
    \begin{itemize}
    \item[$\circ$] embarrassing: the mathematical problem is a DAE
    \item[$\circ$] time step size is determined by the coupled solution velocity $\bU$
         \begin{itemize}
         \item[$\vartriangleright$] CFL
         \end{itemize}
    \end{itemize}
\item however, the performance/usability/accuracy consequences of suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
\end{itemize}
\end{frame}

\begin{frame}{time-stepping in future ice sheet models}

\begin{itemize}
\item \alert{implicit time-stepping} should be used for DAE problems
\item a sequence of difficult free-boundary problems must be solved at each time step: \alert{NCP coupled to Stokes}
    \begin{itemize}
    \item[$\circ$] not really ``equations'', because of $s\ge b$ inequality
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item whether explicit or implicit, a non-shallow model solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
\item how fast is the numerical solution process?
    \begin{itemize}
    \item[$\circ$] how do solution algorithms \alert{scale with increasing resolution}?
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item consider the 2D Poisson equation:
    $$-\grad^2 u = f \text{ in $\Omega$}, \quad u=0 \text{ on $\partial\Omega$}$$
\item discretization generates a linear system $A\,\bu = \bb$
\item data size $m$ = (number of unknowns)
    \begin{itemize}
    \item[$\circ$] for low-order discretizations, $m$ = (number of nodes in the grid)
    \item[$\circ$] $m$ scales with mesh cell diameter $h$: \quad $m \sim h^{-2}$ in 2D
    \end{itemize}
\item \alert{complexity} or \alert{algorithmic scaling} of flops, as $m\to\infty$, depends on solver algorithm:
    \begin{itemize}
    \item[$\circ$] $O(m^3)$ for direct linear algebra, ignoring matrix structure
    \item[$\circ$] $\approx O(m^2)$ for sparsity-exploiting direct linear algebra
    \item[$\circ$] $O(m^1)$, \alert{optimal}, for \alert{multigrid} solvers (below)
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item $m$ = (number of velocity and pressure unknowns)
\item I'll parameterize scaling as $O(m^{1+\alpha})$, with $\alpha=0$ optimal
\item \alert{near-optimal solvers} already exist: \hfill $\leftarrow$ \emph{good news!}
    \begin{itemize}
    \item[$\circ$] $\alpha=0.08$ for Isaac et al.~(2015) Stokes solver
        \begin{itemize}
        \item[$\vartriangleright$] unstructured quadrilateral/tetrahedral mesh, $Q_k\times Q_{k-2}$ stable elements, Schur-preconditioned Newton-Krylov, ice-column-oriented algebraic multigrid (AMG) preconditioner for $(\bu,\bu)$ block
        \end{itemize}

\begin{center}
\includegraphics[width=0.6\textwidth]{images/isaac-antarctica.png}
\end{center}
    \item[$\circ$] $\alpha=0.05$ for Tuminaro et al~(2016) 1st-order (shallow) AMG solver
    \item[$\circ$] similar for Brown et al~(2013) 1st-order GMG solver
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{the analysis set-up: definitions and relationships}

\begin{itemize}
\item ice sheets are thin layers, thus ice sheet models often have $O(1)$ mesh points in the vertical direction
    \begin{itemize}
    \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
    \item[$\circ$] \emph{message}: I am ignoring refinement in the vertical
    \end{itemize}

\medskip
\item let $m$ = (\# surface elev, velocity, \& pressure unknowns)
\item for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
  $$m \sim \frac{L^2}{h^2} \hspace{20mm}$$

@@

\begin{frame}{simplified analysis of ice sheet model performance}

\begin{itemize}
\item for a time-stepping model our metric is \alert{flops per model year}
    \begin{itemize}
    \item[$\circ$] glaciologists want high-resolution simulations of ice sheets over 100k year ice age cycles
    \end{itemize}

@@

\bigskip
\begin{center}
\begin{minipage}{0.82\textwidth}
how does flops per model year \alert{scale} in the \alert{high spatial resolution limit} $h\to 0$, equivalently $m\to \infty$?
\end{minipage}
\end{center}
\end{itemize}
\end{frame}

@@

\bigskip
\begin{itemize}
\item \alert{explicit} time-stepping \alert{implies many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] stress-balance solver scaling $\alpha$ is important, but even optimality ($\alpha=0$) cannot fix performance
    \end{itemize}
\end{itemize}
\end{frame}

>>>>

\newcommand{\bb}{\mathbf{b}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}

@@

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$  is defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice
    \end{itemize}
\item<3> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which (solution) velocity $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

@@

\medskip
\item conservation of mass is important both
    \begin{itemize}
    \item[$\circ$] in the icy domain $\Lambda(t) \subset \RR^3$:
\begin{align*}
\text{\alert{incompressibility}}&& \Div \bu &= 0 \qquad \text{in } \Lambda(t), \\
    \intertext{\item[$\circ$] and on the ice surfaces:}
\text{\alert{surface kinematic equation (SKE)}}&& \frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a \qquad \text{on } \Gamma_s(t), \\
\text{\alert{non-penetration}}&&     \bu|_b \cdot \bn_b &= 0 \qquad \text{on } \Gamma_b(t).
\end{align*}

@@

\bigskip
\item {\footnotesize for more on this perspective see Bueler (2021)}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \only<1>{\alert{nonlinear complementarity problem} (NCP)} \only<2>{nonlinear complementarity problem (NCP) coupled to \alert{Stokes}}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t)$}}
\end{align*}

@@

\begin{itemize}
\item for this slide, forget complementarity and boundary conditions
\item result: SKE coupled to Stokes
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

\item only the first of these 5 equations has a time derivative
    \begin{itemize}
    \item[$\circ$] because ice is very viscous and incompressible
    \end{itemize}
\item<2> this time-dependent problem is a \alert{differential algebraic equation} (DAE), an extremely stiff system:
\begin{align*}
\dot x &= f(x,y) \\
     0 &= g(x,y)
\end{align*}

    \begin{itemize}
    \item<2>[$\circ$] but in $\infty$ dimensions (PDAE?), and subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] when pressed, most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model?
    \item[$\circ$] well-posedness of the lubrication approximation of the model has been considered; existence proved in (Jouvet \& Bueler 2012)
    \end{itemize}

\medskip
\item numerical modelers tend to think of the Stokes problem separately from surface evolution
    \begin{itemize}
    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is often taken for granted
    \end{itemize}

\medskip
\item ice sheet geometry evolution is addressed with minimal awareness of complementarity

\medskip
\item NCP-coupled-to-Stokes is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item thickness transport form helps for evolution or stability questions
\item define:
\begin{align*}
H(t,x,y) &=s-b && \text{ice thickness} \\
\bU(t,x,y) &= \frac{1}{H} \int_b^s \bu \,dz && \begin{matrix} \text{vertically-averaged} \\ \hspace{-1.8mm} \text{horizontal velocity} \end{matrix}
\end{align*}

    \begin{itemize}
    \item[$\circ$] note $s$ and $H$ are equivalent variables for modeling ice geometry
    \end{itemize}
\item the \alert{mass continuity equation} for thickness follows from SKE and incompressibility:
   $$\frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) = a$$

\smallskip
\item<2> \emph{question:} is this really an advection equation?
\item<2>[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\item<2> NCP-coupled-to-Stokes DAE system has no characteristic curves
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item the diffusion schema is literal in the lubrication approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item but the fact that ice flows downhill has \emph{time-stepping stability} consequences
    \begin{itemize}
    \item[$\circ$] regardless of your preference for the advective schema!
    \end{itemize}
\item note both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)
\item SIA version of the NCP:
{\small
  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
}
the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$

@@

\medskip
\item $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item[$\circ$] \emph{local} in surface and bed topography
    \item[$\circ$] existence is known for this NCP problem (Jouvet \& Bueler, 2012), when written as a \alert{variational inequality} weak form
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item from now on, let us avoid shallowness approximations
\item then the basic ice sheet model (NCP coupled to Stokes) problem has a \alert{non-local} surface velocity function ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$
{\small
\begin{equation*}
s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0
\end{equation*}
}
\item \emph{figure:} the Stokes velocity solution responds to a surface perturbation by up- and down-stream changes, for several ice thicknesses, \onslide<2>{while the SIA velocity responds only underneath the surface perturbation}
\end{itemize}

@@

\begin{minipage}[t][60cm][t]{\textwidth}
\begin{itemize}
\only<1>{
\item let us recall some traditional numerical analysis
}
\only<2>{
\item \alert{explicit} time stepping is common for \alert{advections}
\item for example, forward Euler using spacing $h$ and time step $\Delta t$:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} + \frac{\bq_{j+1/2}^\ell - \bq_{j-1/2}^\ell}{h} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] need good approximations of flux $\bq=\bU H$: upwinding, Lax-Wendroff, streamline diffusion, flux-limiters, \dots
    \item[$\circ$] conditionally stable, with CFL maximum time step
        $$\Delta t \le \frac{h}{\max |\bU|} = O(h)$$
    \end{itemize}
}
\only<3>{
\item \alert{explicit} time stepping for \alert{diffusions} is best avoided
\item for example, forward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell} + s_{j}^{\ell}) - D_{j-\frac{1}{2}} (s_{j}^{\ell} + s_{j-1}^{\ell})}{h^2} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] conditionally stable, with maximum time step
        $$\Delta t \le \frac{h^2}{\max D} = O(h^2)$$
    \end{itemize}
}
\only<4>{
\item \alert{implicit} time stepping for \alert{diffusions} is often recommended
\item for example, backward Euler:
	$$\frac{H_j^{\ell+1} - H_j^\ell}{\Delta t} - \frac{D_{j+\frac{1}{2}} (s_{j+1}^{\ell+1} + s_{j}^{\ell+1}) - D_{j-\frac{1}{2}} (s_{j}^{\ell+1} + s_{j-1}^{\ell+1})}{h^2} = a_j^\ell$$

    \begin{itemize}
    \item[$\circ$] unconditionally stable, but must solve equations at each step
    \item[$\circ$] further implicit schemes: Crank-Nicolson, BDF, \dots
    \end{itemize}
}
\end{itemize}
\end{minipage}
\end{frame}

\begin{frame}{time-stepping in current and future ice sheet models}

\begin{itemize}
\item current-technology large-scale models use explicit time stepping
    \begin{itemize}
    \item[$\circ$] this is embarrassing: the mathematical problem is a DAE
    \item[$\circ$] the accuracy/performance/usability consequences of the suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
    \end{itemize}
\item most researchers believe the advection schema
    \begin{itemize}
    \item[$\circ$] time step is determined by CFL using coupled solution velocity $\bU$
    \end{itemize}

\medskip
\item \alert{implicit time-stepping} is appropriate for DAE problems
\item a sequence of NCP-coupled-to-Stokes free-boundary problems must be solved at each time step
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item a non-shallow model solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
\item how fast is the numerical solution process?
    \begin{itemize}
    \item[$\circ$] how do solution algorithms \alert{scale} with increasing spatial resolution?
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item for example, consider the 2D Poisson equation:
    $$-\grad^2 u = f \,\text{ in $\Omega$}, \qquad u=0 \,\text{ on $\partial\Omega$}$$
\item discretization generates a linear system $A\,\bu = \bb$ with $\bu\in\RR^m$
\item data size $m$ is the number of unknowns
    \begin{itemize}
    \item[$\circ$] for low-order discretizations, $m$ = \#(nodes in the grid)
    \item[$\circ$] $m$ scales with mesh cell diameter $h$: \quad $m \sim h^{-2}$ in 2D
    \end{itemize}
\item \alert{complexity} or \alert{algorithmic scaling} of flops, as $m\to\infty$, depends on solver algorithm:
    \begin{itemize}
    \item[$\circ$] $O(m^3)$ for direct linear algebra, ignoring matrix structure
    \item[$\circ$] $\approx O(m^2)$ for sparsity-exploiting direct linear algebra
    \item[$\circ$] $O(m^1)$, \alert{optimal}, e.g.~for \alert{multigrid} solvers (below)
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item Stokes: \quad $m$ = \#(velocity and pressure unknowns)
\item model the scaling as $O(m^{1+\alpha})$, with $\alpha=0$ optimal
\item \alert{near-optimal solvers} already exist: \hfill $\leftarrow$ \emph{good news!}
    \begin{itemize}
    \item[$\circ$] $\alpha=0.08$ for Isaac et al.~(2015) Stokes solver
        \begin{itemize}
        \item[$\vartriangleright$] unstructured quadrilateral/tetrahedral mesh, $Q_k\times Q_{k-2}$ stable elements, Schur-preconditioned Newton-Krylov, ice-column-oriented algebraic multigrid (AMG) preconditioner for $(\bu,\bu)$ block
        \end{itemize}

\begin{center}
\includegraphics[width=0.6\textwidth]{images/isaac-antarctica.png}
\end{center}
    \item[$\circ$] $\alpha=0.05$ for Tuminaro et al~(2016) 1st-order (shallow) AMG solver
    \item[$\circ$] similar for Brown et al~(2013) 1st-order (shallow) GMG solver
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{the analysis set-up}

\begin{itemize}
\item ice sheets are thin layers, thus ice sheet models often have $O(1)$ mesh points in the vertical direction
    \begin{itemize}
    \item[$\circ$] e.g.~Issac et al (2015) Stokes solver
    \item[$\circ$] \emph{simple message}: I am ignoring refinement in the vertical
    \end{itemize}

\medskip
\item let $m$ = \#(surface elevation \& velocity \& pressure unknowns)
\item for map-plane domain $\Omega \subset \RR^2$ of width $L$ and cells of diameter $h$:
  $$m \sim \frac{L^2}{h^2} \hspace{20mm}$$

@@

\begin{frame}{the simplified ice sheet model performance question}

\begin{itemize}
\item glaciologists want to run time-stepping high-resolution simulations of ice sheets over e.g.~$10^5$ year ice age cycles

\bigskip
\item proposed metric: \quad \alert{flops per model year}

@@

\bigskip
\begin{center}
\begin{minipage}{0.82\textwidth}
how does this metric \alert{scale} in the \alert{high spatial resolution limit} $h\to 0$, equivalently $m\to \infty$?
\end{minipage}
\end{center}
\end{itemize}
\end{frame}

@@

\bigskip
\begin{itemize}
\item \alert{explicit} time-stepping \alert{implies many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] stress-balance solver scaling $\alpha$ is important,
    \item[$\circ$] but even optimality ($\alpha=0$) cannot rescue performance
    \end{itemize}
\end{itemize}
\end{frame}
